\textsf{Are large language models superhuman chemists?}

### 2024-03-21 src/tex/ms.tex
typesetting

    We curated more than 6000 question answer pairs about a wide array of subfields, evaluated leading models such as GPT-4, and found that the best models, on average, outperform the best human chemists in our study. 
    The models, however, struggle with some chemical reasoning tasks that are easy for human experts and give overconfident misleading predictions, e.g., about the safety profile of chemicals.

@@

Thus, to better understand what \glspl{llm} can do for chemistry and materials science, and where they might be improved with further developments, evaluation frameworks are needed to allow us to systematically measure progress and mitigate potential harms.
For the development of \glspl{llm}, evaluation is currently mostly performed via standardized benchmarks suites such as BigBench\autocite{srivastava2022beyond} or the LM Eval Harness.\autocite{eval-harness}
The former contains, among 204 tasks (such as linguistic puzzles), only two tasks classified as \enquote{chemistry related} whereas the latter contains no specific chemistry tasks.
Due to the lack of widely excepted standard benchmarks, the developers of chemical language models\autocite{jablonka2024leveraging, guo2023large, ahmad2022chemberta2, Cai_2024, frey2023neural} frequently utilize language-interfaced\autocite{dinh2022lift} tabular datasets such as the ones reported in MoleculeNet,\autocite{wu2018moleculenet} Therapeutic Data Commons\autocite{huang2021therapeutics} or MatBench.\autocite{dunn2020benchmarking}
In these cases, the models are  evaluated on predicting very specific properties of molecules (e.g., solubility, toxicity, or reactivity) or on predicting the outcome of specific chemical reactions.
This, however, only gives a very limited view of the general chemical capabilities of the models.

@@

\begin{figure}
    \centering
    \includegraphics{figures/question_count_barplot.pdf}
    \caption{\textbf{Number of questions for different topics.} The topics have been assigned using a combination of a rule-based system (mostly based on the source the question has been sampled from) as well as a classifier operating on a word-embedding of the question. 
    The figure shows that not all aspects of chemistry are equally represented in our corpus. The \chembench corpus, by design, currently focuses on safety-related aspects, which is also evident in \Cref{fig:question_diversity}. This figure represents the combined count of \glstext{mcq} and open-ended questions.}
    \label{fig:topic_barplot}
    \script{plot_statistics.py}
\end{figure}

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{figures/question_diversity.pdf}
    \caption{\textbf{Principal component projection of embeddings of questions in the \chembench corpus.} To obtain this figure, we embedded question and answers using the BART model\autocite{bart} (using other embeddings, such as of OpenAI's ada model leads to qualitatively similar results). We then project the embeddings into a two-dimensional space using \gls{pca}. We color the points based on a classification into topics. Safety related aspects cover a large part of the figure that is not covered by questions from other topics.}
    \label{fig:question_diversity}
    \script{plot_question_diversity.py}
\end{figure}

While many of the existing benchmarks are designed around \gls{mcq}, this does not reflect the reality of chemistry education and research.
For this reason, \chembench samples both \gls{mcq} and open-ended questions (\variable{output/mcq_questions.txt} \gls{mcq} questions vs. \variable{output/non_mcq_questions.txt} open-ended questions).
%As one might expect, most questions are difficult to read according to the Flesch–Kincaid readability test (\variable{output/flesch_kincaid_reading_ease.txt}).\autocite{kincaid1975derivation}

@@

% emphasize that the topics are more balanced in this subset

@@

Because the text used in scientific settings is different from typical natural language, many models have been developed that deal with such text in a special way.
For instance, Galactica\autocite{taylor2022galactica} uses special tokenization or encoding procedures for molecules and equations. 
Current benchmarking suites, however, do not account for such special treatment of scientific information.
To address this, \chembench is designed to encode the semantic meaning of various parts of the question or answer.  
For instance, molecules represented in \gls{smiles} are encapsulated in \texttt{[START\_SMILES][\textbackslash END\_SMILES]}.

@@

To understand the current capabilities of \glspl{llm} in the chemical sciences, we evaluated a wide range of leading models\autocite{Huggingface} on the \chembench corpus, including systems augmented with external tools.
An overview of the results of this evaluation is shown in \Cref{fig:human_vs_models_bar}. 
In this figure, we show the percentage of questions that the models answered completely correctly.
Moreover, we show the worst, best, and average performance of the human experts in our study which we obtained via a custom web application (\url{chembench.org}) that we used to survey the experts.
Remarkably, the figure shows that the leading \gls{llm}, GPT-4, outperforms the best human in our study in this overall metric and vastly, by more than a factor of two, exceeds the average performance of the experts in our study.
Many other models also outperform the average human performance. It is, however, interesting to note that the Galactica model that was trained specifically for scientific applications is outperformed by more general models.
% Bar chart

\begin{figure}[htb]
    \centering
    \includegraphics{figures/human_subset_performance.pdf}
    \caption{\textbf{Performance of models and humans on the \enquote{tiny} subset of the  \chembench corpus.} The figure shows the percentage of questions that the models answered completely correctly. We use horizontal bars to indicate the performance of various models and highlight statistics of the human performance. 
    Since the humans did not answer all the questions, this plot is based on the subset of questions that most humans answered.
    The metric we use here is harsh as it only considers a question answered completely correctly or completely incorrectly, partially correct answers are also considered as incorrect.
    \Cref{fig:barplot_all_correct_all_questions} gives an overview of the performance of various models on the entire corpus.
    }
    \label{fig:human_vs_models_bar}
    \script{plot_overview_performance_plot.py}
\end{figure}

Given the large interest in tool-augmented systems, the bad performance of these systems in our benchmark is striking. 
Their lack of performance, however, is partially due to the fact that we limited the system to a maximum of ten \gls{llm} calls.
With the default tool augmented setup (using the so-called ReAct method\autocite{yao2023react}), however, this often did not allow the system to find the correct answer (e.g., because it repeatedly tried to search for the answer on the web and did not find a solution within ten calls to the \gls{llm}).
This observation highlights the importance of communicating and measuring not only in predictive performance but also computational cost (e.g., in terms of \gls{api} calls) for tool-augmented systems.

%% Note also while the performance of the systems is better than what humans do, it is still far from perfect

% Radar chart
\paragraph{Performance per topic} To obtain a more detailed understanding of the performance of the models, we also analyzed the performance of the models in different subfields of the chemical sciences.
For this analysis, we defined a set of topics (see \Cref{sec:meth-topic}) and classified the questions into these topics based and hand-crafted rules as well as classifier models.
We then computed the percentage of questions that the models, or humans, answered completely correctly for each topic.

\begin{figure}[htb]
    \centering
    \includegraphics{figures/all_questions_models_completely_correct_radar_human.pdf}
    \caption{\textbf{Performance of the models and humans on the different topics of the \enquote{tiny} subset of the \chembench corpus.} The radar plot shows the performance of the models and humans on the different topics of \enquote{tiny} subset of the \chembench corpus. The performance is measured as the fraction of questions that were answered completely correctly by the models.
    The best score for every dimension is one (all questions answered correctly) and the worst score is zero (no question answered completely correctly). A larger colored area indicates a better performance.
    This figure shows the performance on the subset of questions that was answered by humans, too. The performance of models on the entire corpus is shown in \Cref{fig:performance_per_topic}.
    }
    \label{fig:all_questions_models_completely_correct_radar_human}
    \script{analyze_model_reports.py}
\end{figure}

In this spider chart, the worst score for every dimension is zero (no question answered completely correctly) and the best score is one (all questions answered correctly). 
Thus, a larger colored area indicates a better performance. 
One can observe that this performance varies widely across models and topics. 
While macromolecular chemistry and biochemistry receive relatively high scores for many models, this is not the case for topics such as chemical safety or analytical chemistry.
In the subfield of analytical chemistry the prediction of the number of peaks observable in a \gls{nmr} spectrum proved difficult for the models (e.g., \variable{output/subset_scores/is_number_nmr_peaks_gpt4.txt} percent correct answers for GPT-4) while this question appeared easier (\variable{output/human_subset_scores/is_number_nmr_peaks.txt} percent correct) for trained humans.
Importantly, the human experts are given a drawing of the compounds, whereas models are only shown the \gls{smiles} string of a compound and have to use this to reason about the symmetry of the compound (i.e., to identify the number of diasterotopically distinct protons, which requires \emph{reasoning} about the topology and structure of a molecule). 
This is mirrored in the questions about the number of isomers, which is models struggle with more than humans (\variable{output/subset_scores/is_number_of_isomers_gpt4.txt} percent for GPT-4 vs. \variable{output/human_subset_scores/is_number_of_isomers.txt} percent for humans).
These findings also shine an interesting light on the value of textbook-inspired questions. 
A subset of the questions in the \chembench are based on textbooks targeted at undergraduate students. 
On those questions the models tend to perform better than on some our semi-automatically constructed tasks (see \Cref{fig:performance_per_topic}).
We suppose that this is due to the fact that some of the textbook questions require less reasoning and more memorization and combination of facts.

% GFK performance -> would the models obtain a chemical license 
% Interesting twist perhaps: the performance on the university textbooks or exams is better than on the ones we automatically constructed
% in contrast to LLM the human performance is also less spiky
% emphasise that the exact ranking depends on the subset of questions we look at

@@

@@

\begin{figure}[htb]
    \centering
    \includegraphics{figures/confidence_vs_performance_overall.pdf}
    \caption{\textbf{Relationship between confidence of the model in the answer and correctness.} For this analysis, we used verbalized confidence estimates from the model. To obtain those estimates, we prompted the models to return a confidence score on an ordinal scale. We then plot the correctness of the answers [which can be either completely correct (1) or completely incorrect (0)] against the confidence score.
        The stripplots show the individual data points, a strong density indicating a higher number of points. The lines show the mean and the errorbars show the standard error of the mean. The figure shows that the reliability of the confidence estimates varies widely across models. While the ones for GPT-4 and Claude-2 are not particularly reliable, the ones for Claude-3 follow the expected trend.}
    \label{fig:confidence_vs_performance}
    \script{joint_analysis_confidence_performance.py}
\end{figure}

@@

Our findings also highlight the nuanced tradeoff between breadth and depth of evaluation frameworks. % link to the missing correlation 
% also need for better ways of how humans actually use LLMs. How do they get the confidence?  
% an LLM is not a chemist and at least the current systems seem to be far from reasoning like a chemist 
% for fact like questions, that requires knowledge they might have seen on the internet, they work well, but for reasoning they are not good yet.

The current evaluation frameworks for \glspl{llm} are mostly designed to measure the performance of the models on specific tasks. 
They cannot easily be used to evaluate systems, or models built for scientific applications. Thus, we had on little understanding of the capabilities of \glspl{llm} in the chemical sciences.
Our work,  shows that carefully curated benchmarks can provide a more nuanced understanding of the capabilities of \glspl{llm} in the chemical sciences.
It lays the groundwork for the development of more advanced systems that can be used in the chemical sciences and highlights the importance of continued development of evaluation frameworks for systematic improvement of safe and useful \glspl{llm}.
Clearly defined metrics have been the key for the progress of many fields of \gls{ml} such computer vision. 
We hope that our open source \chembench will be a stepping stone for the development of such breakthrough for the chemical sciences.

@@

To ensure that the questions do not enter a training dataset we use the canary string of the Big Bench project.
This requires that \Gls{llm} developers filter their training dataset for this canary string.\autocite{openai2024gpt4, srivastava2022beyond}

\begin{figure}[htb]
    \includegraphics[width = \textwidth]{figures/chem-bench.pdf}
    \caption{\textbf{Overview of the workflow for the assembly of the \chembench corpus}. To assemble the \chembench corpus, we first collected questions from various sources. Some tasks were manually curated, others semi-programmatically. For all questions we added semantic annotations to make them compatible with systems that use special processing for modalities that are not conventional natural text. Before adding the questions to the corpus, we reviewed with manual as well as automatic methods.}
    \label{fig:curation_workflow}
\end{figure}

@@

\subparagraph{Organic reactivity}
\variable{output/question_count_per_dir/json_file_counts_organic_reactivity.txt}  questions are primarily sourced from tutorials and exam papers at for the Bachelor of Science in Chemistry at the Technical University of Munich (Germany), exploring reactions and mechanisms in organic chemistry.

@@

\subparagraph{Chemical safety} 
This category comprises a total of \variable{output/question_count_per_dir/json_file_counts_safety.txt} questions, including semi-programmatically created questions covering areas such as GHS classification, hazard statements, \gls{dai}, and pictograms from the PubChem database \autocite{pubchem} (see below for details). 
Additionally, it explores crucial topics like materials' compatibility (\variable{output/question_count_per_dir/json_file_counts_materials_compatibility.txt} questions) and chemical compatibility (\variable{output/question_count_per_dir/json_file_counts_chem_chem_comp.txt} questions), essential for understanding the hazardous impact of mixing certain materials and chemicals. 
Furthermore, it includes \variable{output/question_count_per_dir/json_file_counts_blac_gfk.txt} questions based on the question bank published by the Federal/State Working Group on Chemical Safety (BLAC), which is used for the expertise examination (\enquote{Sachkundeprüfung}) according to §11 of the Chemical Prohibition Ordinance (\enquote{Chemikalien-Verbotsverordnung}).
Moreover, this category addresses questions lab safety and topics such as pharmacology, toxicology, material safety data sheets, and chemical safety data sheets. 
The chemical safety category stands out as one of the primary focus areas within the \chembench corpus.

@@

To maintain the consistency with the training, we employ distinct prompt templates in general tailored for completion models and instruction-tuned models. 
We impose constraints on the models within these templates to receive response in a specific format so that a robust, fair and consistent parsing can be performed as explained below.
Certain models are trained with special annotations and \LaTeX\xspace syntax for scientific notations, chemical reactions, or symbols embedded within the text. 
For example, all the SMILES representations are encapsulated within \texttt{[START\_SMILES][\textbackslash END\_SMILES]} in Galactica\autocite{taylor2022galactica}.
Our prompting strategy consistently adheres to  these details in a model-specific manner by post-processing \LaTeX\xspace syntax, chemical symbols, chemical equations and physical units (by either adding or removing wrappers).
This step can be easily customized in our codebase.

\paragraph{Parsing}
Our parsing workflow is, by default, multistep and primarily based on regular expressions.
In the case of instruction-tuned models we first attempt to identify the \texttt{[ANSWER][\textbackslash ANSWER]} environment we prompt the model to report the answer in.
In the case of completion models, this step is skipped. From there, we attempt to extract the relevant enumeration letters (for multiple-choice questions) or numbers.
In the case of numbers, our regular expression was engineered to be able to deal with various forms of scientific notation.
As initial tests indicated that models sometimes return integers in the form of words, e.g. \enquote{one} instead of \enquote{1}, we also implemented a word-to-number conversion using regular expressions.
In case these hard-coded parsing steps fail, we fall back to using a \gls{llm}, e.g. Claude-2, to parse the completion.
%Manual verification indicates that this \gls{llm}-based parsing is not a relevant error source and has been performed correctly in all cases we checked (randomly sampled 4 questions per topic).

@@

\subparagraph{Tool augmented models}
In addition to directly prompting \glspl{llm}, we also investigated the performance of tool-augmented systems.
For this we, on the one hand, investigated the 7B parameter \enquote{online} model of Perplexity.AI and, on the other hand, gpt-3.5-turbo as well as Claude 2, with ReAct-style tool augmentation.\autocite{yao2023react}
The latter two models had access to WolframAlpha, the ArXiv \gls{api}, a Python interpreter, as well as web search (using DuckDuckGo).
We implemented the system using Langchain with the default prompts and constrained the system to a maximum of ten \gls{llm} calls.

@@

\paragraph{Comparison with models}
To compare the performance of humans (who might have answered only some questions) with the performance of models (which answered all questions), we focussed on questions which at least four humans answered and limited the pool of human scorers to those who answered at least 100 questions (i.e., \variable{output/num_humans_with_more_than_100_scores.txt} humans). 
The latter threshold was chosen to limit it to humans who made serious attempts at systematically answering a part of the questions. 
This analysis might lead to potential biases, most likely in favor of humans as they were allowed to skip questions. \variable{output/num_humans_with_204_scores.txt} humans answered more than 200 questions.

\subsection{Classification of questions into topics}\label{sec:meth-topic} When curating our dataset we systematically recorded keywords and sources.
To allow for analysis of the model performance as a function of the topic, we leverage this information together with sequence classification models.
For questions which can easily be assigned to a topic based on the source (e.g., number of \gls{nmr} peaks, chemical compatibility, toxicology exam questions) we use this information to make the assignment.
For the remaining ones, e.g., from chemistry olympiad questions, we use zero-short sequence classification\autocite{zeroshotsequence} using the BART model\autocite{bart, FacebookBART}, which our preliminary analysis found to be more robust than topic modeling based on embeddings from OpenAI's \texttt{ada} model or Cohere's \texttt{Cohembed-english-v3.0} model.

@@

\section*{Acknowledgements}
This work was supported by the Carl Zeiss Foundation and a \enquote{Talent Fund} of the \enquote{Life} profile line of the Friedrich Schiller University Jena.
We also want to acknowledge access to the HPC cluster of Stability AI.
M.A. expresses gratitude to the European Research Council (ERC) for evaluating the project with the reference number 101106377 titled \enquote{CLARIFIER} and accepting it for funding under the HORIZON TMA MSCA Postdoctoral Fellowships - European Fellowships. 
Furthermore, M.A. acknowledges the funding provided by UK Research and Innovation (UKRI) under the UK government’s Horizon Europe funding guarantee (Grant Reference: EP/Y023447/1; Organization Reference: 101106377).

@@

\section*{Conflicts of interest}
K.M.J.\ is a paid consultant for OpenAI (as part of the red teaming network). M.P.\ is employee of Stability AI and A.M.\ and N.A.\ are paid contractors of Stability AI.

@@

\footnotesize
\insertcredits
\normalsize
\printbibliography
\end{refsection}

>>>>

    We curated more than 6000 question answer pairs about a wide array of subfields, evaluated leading models such as GPT-4 and Claude 3, and found that the best models, on average, outperform the best human chemists in our study. 
    The models, however, struggle with some chemical reasoning tasks that are easy for human experts and give overconfident misleading predictions, e.g., about the safety profile of chemicals.

@@

Thus, to better understand what \glspl{llm} can do for chemistry and materials science, and where they might be improved with further developments, evaluation frameworks are needed to allow us to systematically measure progress and mitigate potential harms.
For the development of \glspl{llm}, evaluation is currently mostly performed via standardized benchmarks suites such as BigBench\autocite{srivastava2022beyond} or the LM Eval Harness.\autocite{eval-harness}
The former contains, among 204 tasks (such as linguistic puzzles), only two tasks classified as \enquote{chemistry related} whereas the latter contains no specific chemistry tasks.
Due to the lack of widely accepted standard benchmarks, the developers of chemical language models\autocite{jablonka2024leveraging, guo2023large, ahmad2022chemberta2, Cai_2024, frey2023neural} frequently utilize language-interfaced\autocite{dinh2022lift} tabular datasets such as the ones reported in MoleculeNet,\autocite{wu2018moleculenet} Therapeutic Data Commons\autocite{huang2021therapeutics} or MatBench.\autocite{dunn2020benchmarking}
In these cases, the models are  evaluated on predicting very specific properties of molecules (e.g., solubility, toxicity, or reactivity) or on predicting the outcome of specific chemical reactions.
This, however, only gives a very limited view of the general chemical capabilities of the models.

@@

\begin{figure}[!htb]
    \centering
    \includegraphics{figures/question_count_barplot.pdf}
    \caption{\textbf{Number of questions for different topics.} The topics have been assigned using a combination of a rule-based system (mostly based on the source the question has been sampled from) as well as a classifier operating on word-embeddings of the questions. 
    The figure shows that not all aspects of chemistry are equally represented in our corpus. The \chembench corpus, by design, currently focuses on safety-related aspects, which is also evident in \Cref{fig:question_diversity}. This figure represents the combined count of \glstext{mcq} and open-ended questions.}
    \label{fig:topic_barplot}
    \script{plot_statistics.py}
\end{figure}

\begin{figure}[!htb]
    \centering
    \includegraphics[width=\textwidth]{figures/question_diversity.pdf}
    \caption{\textbf{Principal component projection of embeddings of questions in the \chembench corpus.} To obtain this figure, we embedded questions and answers using the BART model\autocite{bart} (using other embeddings, such as of OpenAI's ada model leads to qualitatively similar results). We then project the embeddings into a two-dimensional space using \gls{pca}. We color the points based on a classification into topics. Safety related aspects cover a large part of the figure that is not covered by questions from other topics.}
    \label{fig:question_diversity}
    \script{plot_question_diversity.py}
\end{figure}

While many of the existing benchmarks are designed around \gls{mcq}, this does not reflect the reality of chemistry education and research.
For this reason, \chembench samples both \gls{mcq} and open-ended questions (\variable{output/mcq_questions.txt} \gls{mcq} questions vs. \variable{output/non_mcq_questions.txt} open-ended questions).

@@

@@

\paragraph{Benchmark suite design} Because the text used in scientific settings is different from typical natural language, many models have been developed that deal with such text in a special way.
For instance, Galactica\autocite{taylor2022galactica} uses special tokenization or encoding procedures for molecules and equations. 
Current benchmarking suites, however, do not account for such special treatment of scientific information.
To address this, \chembench is designed to encode the semantic meaning of various parts of the question or answer.  
For instance, molecules represented in \gls{smiles} are encapsulated in \texttt{[START\_SMILES][\textbackslash END\_SMILES]} tags. 
This allows the model to treat the \gls{smiles} string different from other text. 
\chembench can seamlessly handle such special treatment in an easily extensible way because the questions are stored in an annotated format.

@@

\begin{figure}[!h]
    \centering
    \includegraphics{figures/human_subset_performance.pdf}
    \caption{\textbf{Performance of models and humans on the \enquote{tiny} subset of the  \chembench corpus.} The figure shows the percentage of questions that the models answered completely correctly. We use horizontal bars to indicate the performance of various models and highlight statistics of the human performance. 
    Since the humans did not answer all the questions, this plot is based on the subset of questions that most humans answered.
    The evaluation we use here is very strict as it only considers a question answered completely correctly or completely incorrectly, partially correct answers are also considered as incorrect.
    \Cref{fig:barplot_all_correct_all_questions} gives an overview of the performance of various models on the entire corpus.
    Systems with \enquote{ReAct} in the name are tool augmented, i.e., they can call external tools such as web search or Python code executors to better answer the questions.
    However, we limit those systems to a maximum of ten calls to the \gls{llm}. This constraint led the systems to often not find the correct answer within the specified number of calls.
    In this case, we consider the answer as incorrect. 
    }
    \label{fig:human_vs_models_bar}
    \script{plot_overview_performance_plot.py}
\end{figure}

\paragraph{System performance} 
To understand the current capabilities of \glspl{llm} in the chemical sciences, we evaluated a wide range of leading models\autocite{Huggingface} on the \chembench corpus, including systems augmented with external tools.
An overview of the results of this evaluation is shown in \Cref{fig:human_vs_models_bar}. 
In this figure, we show the percentage of questions that the models answered completely correctly.
Moreover, we show the worst, best, and average performance of the human experts in our study which we obtained via a custom web application (\url{chembench.org}) that we used to survey the experts.
Remarkably, the figure shows that the leading \gls{llm}, Claude 3, outperforms the best human in our study in this overall metric and vastly, by more than a factor of two, exceeds the average performance of the experts in our study.
Many other models also outperform the average human performance. It is, however, interesting to note that the Galactica model that was trained specifically for scientific applications is outperformed by more general models.

Given the large interest in tool-augmented systems, the bad performance of these systems (GPT-3.5 and Claude 2 augmented with tools) in our benchmark is striking. 
Their lack of performance, however, is partially due to the fact that we limited the system to a maximum of ten \gls{llm} calls.
With the default tool augmented setup (using the so-called ReAct method\autocite{yao2023react}), however, this often did not allow the system to find the correct answer (e.g., because it repeatedly tried to search for the answer on the web and did not find a solution within ten calls to the \gls{llm}).
This observation highlights the importance of communicating and measuring not only in predictive performance but also computational cost (e.g., in terms of \gls{api} calls) for tool-augmented systems.

% Radar chart
\paragraph{Performance per topic} To obtain a more detailed understanding of the performance of the models, we also analyzed the performance of the models in different subfields of the chemical sciences.
For this analysis, we defined a set of topics (see \Cref{sec:meth-topic}) and classified all questions in the \chembench corpus into these topics based on hand-crafted rules as well as classifier models operating on the question texts.
We then computed the percentage of questions that the models, or humans, answered completely correctly for each topic.
\begin{figure}[!h]
    \centering
    \includegraphics{figures/all_questions_models_completely_correct_radar_human.pdf}
    \caption{\textbf{Performance of the models and humans on the different topics of the \enquote{tiny} subset of the \chembench corpus.} The radar plot shows the performance of the models and humans on the different topics of \enquote{tiny} subset of the \chembench corpus. The performance is measured as the fraction of questions that were answered completely correct by the models.
    The best score for every dimension is one (all questions answered correctly) and the worst score is zero (no question answered completely correctly). A larger colored area indicates a better performance.
    This figure shows the performance on the subset of questions that was answered by humans. The performance of models on the entire corpus is shown in \Cref{fig:performance_per_topic}.
    }
    \label{fig:all_questions_models_completely_correct_radar_human}
    \script{analyze_model_reports.py}
\end{figure}
In this spider chart, the worst score for every dimension is zero (no question answered completely correctly) and the best score is one (all questions answered correctly). 
Thus, a larger colored area indicates a better performance. 
One can observe that this performance varies widely across models and topics. 
While macromolecular chemistry and biochemistry receive relatively high scores for many models, this is not the case for topics such as chemical safety or analytical chemistry.
In the subfield of analytical chemistry the prediction of the number of peaks observable in a \gls{nmr} spectrum proved difficult for the models (e.g., \variable{output/subset_scores/is_number_nmr_peaks_gpt4.txt} percent correct answers for GPT-4) while this question appeared easier (\variable{output/human_subset_scores/is_number_nmr_peaks.txt} percent correct) for trained humans.
Importantly, the human experts are given a drawing of the compounds, whereas models are only shown the \gls{smiles} string of a compound and have to use this to reason about the symmetry of the compound (i.e., to identify the number of diasterotopically distinct protons, which requires \emph{reasoning} about the topology and structure of a molecule). 
These findings also shine an interesting light on the value of textbook-inspired questions. 
A subset of the questions in the \chembench are based on textbooks targeted at undergraduate students. 
On those questions the models tend to perform better than on some our semi-automatically constructed tasks (see \Cref{fig:performance_per_topic}).
For instance, while the overall performance in the chemical safety topic is low, the models would pass the examination for a chemical license based on a subset of a questions we sampled from the corresponding question bank (e.g., \variable{output/subset_scores/is_gfk_gpt4.txt}\% correct answers for GPT-4, \variable{output/subset_scores/is_gfk_claude3.txt}\% for Claude 3, and \variable{output/human_subset_scores/is_gfk.txt}\% for the human experts).
While those findings are impacted by the subset of questions we sampled, the results still highlight that good performance on such question bank or textbook questions does not necessarily translate to good performance on other questions that require more reasoning.

@@

\clearpage

@@

\begin{figure}[!h]
    \centering
    \includegraphics{figures/confidence_vs_performance_overall.pdf}
    \caption{\textbf{Relationship between confidence of the model in the answer and correctness.} For this analysis, we used verbalized confidence estimates from the model. To obtain those estimates, we prompted the models to return a confidence score on an ordinal scale. We then plot the correctness of the answers [which can be either completely correct (1) or completely incorrect (0)] against the confidence score.
        The stripplots show the individual data points, a strong density indicating a higher number of points. The lines show the mean and the errorbars show the standard error of the mean. The figure shows that the reliability of the confidence estimates varies widely across models. While the ones for GPT-4 and Claude-2 are not particularly reliable, the ones for Claude-3 follow the expected trend.}
    \label{fig:confidence_vs_performance}
    \script{joint_analysis_confidence_performance.py}
\end{figure}

@@

Our findings also highlight the nuanced tradeoff between breadth and depth of evaluation frameworks. 
The analysis of model performance on different topics shows that the performance of models varies widely across topics. 
But even within a topic, the performance of models can vary widely depending on the type of question and the reasoning required to answer it.

The current evaluation frameworks for \glspl{llm} are mostly designed to measure the performance of the models on specific tasks. 
They cannot easily be used to evaluate systems, or models built for scientific applications. Thus, we had on little understanding of the capabilities of \glspl{llm} in the chemical sciences.
Our work,  shows that carefully curated benchmarks can provide a more nuanced understanding of the capabilities of \glspl{llm} in the chemical sciences.
It lays the groundwork for the development of more advanced systems that can be used in the chemical sciences and highlights the importance of continued development of evaluation frameworks for systematic improvement of safe and useful \glspl{llm}.
Our findings also illustrate that more focus is needed in the development of better human-model interaction frameworks given that models are not able to estimate their own limitations.

While our findings indicate many areas for further improvement of \gls{llm}-based systems, it is also important to realize that clearly defined metrics have been the key for the progress of many fields of \gls{ml} such computer vision. 
We thus believe that although current systems are far from reasoning like a chemist our open source \chembench framework will be a stepping stone for the development of systems that might come closer to this goal.

\clearpage

@@

To ensure that the questions do not enter a training dataset we use the same canary string as the Big Bench project.
This requires that \Gls{llm} developers filter their training dataset for this canary string.\autocite{openai2024gpt4, srivastava2022beyond}

\begin{figure}[!h]
    \includegraphics[width = \textwidth]{figures/chem-bench.pdf}
    \caption{\textbf{Overview of the workflow for the assembly of the \chembench corpus}. To assemble the \chembench corpus, we first collected questions from various sources. Some tasks were manually curated, others semi-programmatically. For all questions we added semantic annotations to make them compatible with systems that use special processing for modalities that are not conventional natural text. Before adding the questions to the corpus, we reviewed with manual as well as automatic methods.}
    \label{fig:curation_workflow}
\end{figure}

@@

\subparagraph{Organic reactivity}
\variable{output/question_count_per_dir/json_file_counts_organic_reactivity.txt}  questions are primarily sourced from tutorials and exam papers for the Bachelor of Science in Chemistry at the Technical University of Munich (Germany), exploring reactions and mechanisms in organic chemistry.

@@

\subparagraph{Chemical safety} 
This category comprises a total of \variable{output/question_count_per_dir/json_file_counts_safety.txt} questions, including semi-programmatically created questions covering areas such as GHS classification, hazard statements, \gls{dai}, and pictograms from the PubChem database \autocite{pubchem} (see below for details). 
Additionally, it explores crucial topics like materials' compatibility (\variable{output/question_count_per_dir/json_file_counts_materials_compatibility.txt} questions) and chemical compatibility (\variable{output/question_count_per_dir/json_file_counts_chem_chem_comp.txt} questions), essential for understanding the hazardous impact of mixing certain materials and chemicals. 
Furthermore, it includes \variable{output/question_count_per_dir/json_file_counts_blac_gfk.txt} questions based on the question bank published by the Federal/State Working Group on Chemical Safety (BLAC), which is used for the expertise examination (\enquote{Sachkundeprüfung}) according to §11 of the Chemical Prohibition Ordinance (\enquote{Chemikalien-Verbotsverordnung}).
Moreover, this category addresses questions about lab safety and topics such as pharmacology, toxicology, material safety data sheets, and chemical safety data sheets. 
The chemical safety category stands out as one of the primary focus areas within the \chembench corpus.

@@

To maintain consistency with the training, we employ distinct prompt templates tailored for completion and instruction-tuned models. 
We impose constraints on the models within these templates to receive response in a specific format so that a robust, fair and consistent parsing can be performed as explained below.
Certain models are trained with special annotations and \LaTeX\xspace syntax for scientific notations, chemical reactions, or symbols embedded within the text. 
For example, all the SMILES representations are encapsulated within \texttt{[START\_SMILES][\textbackslash END\_SMILES]} in Galactica\autocite{taylor2022galactica}.
Our prompting strategy consistently adheres to  these details in a model-specific manner by post-processing \LaTeX\xspace syntax, chemical symbols, chemical equations and physical units (by either adding or removing wrappers).
This step can be easily customized in our codebase.

\paragraph{Parsing}
Our parsing workflow is multistep and primarily based on regular expressions.
In the case of instruction-tuned models we first attempt to identify the \texttt{[ANSWER][\textbackslash ANSWER]} environment we prompt the model to report the answer in.
In the case of completion models, this step is skipped. From there, we attempt to extract the relevant enumeration letters (for multiple-choice questions) or numbers.
In the case of numbers, our regular expression was engineered to be able to deal with various forms of scientific notation.
As initial tests indicated that models sometimes return integers in the form of words, e.g. \enquote{one} instead of \enquote{1}, we also implemented a word-to-number conversion using regular expressions.
In case these hard-coded parsing steps fail, we fall back to using a \gls{llm}, e.g. Claude-2, to parse the completion.
%Manual verification indicates that this \gls{llm}-based parsing is not a relevant error source and has been performed correctly in all cases we checked (randomly sampled 4 questions per topic).

@@

\subparagraph{Tool augmented models}
In addition to directly prompting \glspl{llm}, we also investigated the performance of tool-augmented systems.
For this we investigated the 7B parameter \enquote{online} model of Perplexity.AI. 
In addition, we utilized gpt-3.5-turbo as well as Claude 2, with ReAct-style tool augmentation.\autocite{yao2023react}
The latter two models had access to WolframAlpha, the ArXiv \gls{api}, a Python interpreter, as well as web search (using DuckDuckGo).
We implemented the system using Langchain with the default prompts and constrained the system to a maximum of ten \gls{llm} calls.

@@

\paragraph{Comparison with models}
To compare the performance of humans (who might have answered only some questions) with the performance of models (which answered all questions), we focussed on questions which at least four humans answered and limited the pool of human scorers to those who answered at least 100 questions (i.e., \variable{output/num_humans_with_more_than_100_scores.txt} humans). 
The latter threshold was chosen to limit it to humans who made serious attempts at systematically answering a part of the questions. 
This analysis might lead to potential biases, most likely in favor of humans as they were allowed to skip questions. \variable{output/num_humans_with_204_scores.txt} humans answered more than 200 questions.
For the analysis, we treated each human as a model. For analyses grouped by topic, we computed the topic aggregated averages per human and then averaged over all humans.

\subsection{Classification of questions into topics}\label{sec:meth-topic} When curating our dataset we systematically recorded keywords and sources.
To allow for analysis of the model performance as a function of the topic, we leverage this information together with the output of sequence classification models.
For questions which can easily be assigned to a topic based on the source (e.g., number of \gls{nmr} peaks, chemical compatibility, toxicology exam questions) we use this information to make the assignment.
For the remaining ones, e.g., from chemistry olympiad questions, we use zero-short sequence classification\autocite{zeroshotsequence} using the BART model\autocite{bart, FacebookBART}, which our preliminary analysis found to be more robust than topic modeling based on embeddings from OpenAI's \texttt{ada} model or Cohere's \texttt{Cohembed-english-v3.0} model.

@@

\section*{Acknowledgements}
This work was supported by the Carl Zeiss Foundation and a \enquote{Talent Fund} of the \enquote{Life} profile line of the Friedrich Schiller University Jena.
We thank Stability.AI for the use of its HPC cluster and for its generous donations.
We also thank the OpenBioML.org community and their ChemNLP project team for useful discussions.

M.A. expresses gratitude to the European Research Council (ERC) for evaluating the project with the reference number 101106377 titled \enquote{CLARIFIER} and accepting it for funding under the HORIZON TMA MSCA Postdoctoral Fellowships - European Fellowships. 
Furthermore, M.A. acknowledges the funding provided by UK Research and Innovation (UKRI) under the UK government’s Horizon Europe funding guarantee (Grant Reference: EP/Y023447/1; Organization Reference: 101106377).

@@

\section*{Conflicts of interest}
K.M.J.\ is a paid consultant for OpenAI (as part of the red teaming network). M.P.\ is employee of Stability.AI and A.M.\ and N.A.\ are paid contractors of Stability.AI.

@@

\scriptsize
\insertcredits
\normalsize
\printbibliography
\end{refsection}
